---
title: "Exercise 5"
output: 
  html_document:
    fig_caption: no
    number_sections: no
    toc: no
    toc_float: false
    collapsed: no
    css: html-md-01.css
---

```{r set-options, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=FALSE)
pdf.options(useDingbats = TRUE)
klippy::klippy(position = c('top', 'right'))
```
<p><span style="color: #00cc00;">NOTE:  This page has been revised for Spring 2020, but may undergo further edits.</span></p>
**Geography 4/595:  Geographic Data Analysis**  
**Spring 2020**

**Exercise 5: Some data wrangling and matrix algebra**  
**Finish by Friday, May 1**  

**1. Introduction**

The first aim of this exercise is to illustrate the idea of "data wrangling" or the reshaping or restructuring of input data into the "tidy" form (of a rectangular data set) with variables in columns and observations or cases in rows.  The second part of the exercise consists of a few examples that illustrate the features and application of matrix algebra.

**2. Data and packages**

A the concept of data wrangling, or the reshaping of non-rectangular data set into a rectangular one, can be illustrated using a small sample of monthly climate data for Eugene.  These data are not currently part of the `geog495.RData` workspace file (because they may be read in in different ways--as data frames or "tibbles"), but they can be downloaded here:

- [EugeneClim-short.csv](https://pjbartlein.github.io/GeogDataAnalysis/data/csv/EugeneClim-short.csv) -- Three years (2013-2015) of monthly climate data for Eugene;
- [EugeneClim-short-alt-tvars.csv](https://pjbartlein.github.io/GeogDataAnalysis/data/csv/EugeneClim-short-alt-tvars.csv) -- Temperature data for those three years in an alternative format that is not tidy (i.e. columns are months (or observations) and rows are variables); and
- [EugeneClim-short-alt-pvars.csv](https://pjbartlein.github.io/GeogDataAnalysis/data/csv/EugeneClim-short-alt-pvars.csv) -- Precipitation data, also in an alternative format.

The full data set can be downloaded from here:  [EugeneClim.csv](https://pjbartlein.github.io/GeogDataAnalysis/data/csv/EugeneClim.csv)

(Download these to your current working directory, which can be found using `getwd()`.)

Also install the `tidyverse` package, which in turn installs a number of individual packages that are used in reshaping data.

```{r echo=TRUE, eval=FALSE}
# install the "tidyverse" suite of packages
install.packages("tidyverse")
# library
library(tidyverse)
```

Read in a typical "tidy" `.csv` file, that has variables in columns and observations in rows.    This can be done in the usual way using the `read.csv()` function, which creates a standard data frame, or, if the `reader` packages has been loaded by `library(tidyverse)`, using the `read_csv()` function, which creates a "tibble".  The data here are a "short" three-year-long subset of Eugene monthly climate data.

```{r echo=TRUE, eval=FALSE}
# read a .csv file using the `readr` package
csv_file <- "/Users/bartlein/Documents/geog495/data/csv/EugeneClim-short.csv"
eugclim <- read_csv(csv_file)
eugclim
```

(In the above, you would substitute the path to your working directory.)


Produce a few plots to look at the time series of individual variables, and to look at the annual cycle of each.  The first use of `plot()` below plots the time series of monthly average temperature (`tavg`), while the second illustrates what the annual cycle looks like.

```{r echo=TRUE, eval=FALSE}
# time-series plot
plot(eugclim$tavg ~ eugclim$yrmn, type="o", pch=16, xaxp=c(2013, 2016, 3))
```
```{r echo=TRUE, eval=FALSE}
# by month
plot(eugclim$mon, eugclim$tavg, pch=16, xaxp=c(1, 12, 1))
```

Repeat the plots for some other variables, in particular `prcp` (monthly total precipitation). 

(See (see [https://pjbartlein.github.io/GeogDataAnalysis/lec08.html#variables](https://pjbartlein.github.io/GeogDataAnalysis/lec08.html#variables) for a listing of variables)

> Q1.  Describe the annual cycles of the temperature and moisture-related variables.  When during the year is it colder and when is it warmer, and when is it wetter and when is it drier?

**3. Transforming (reshaping) an alternatively shaped table of data**

An alternative layout for the data table (of just the precipitation-related variables) has the data arranged with variables in rows and months in columns.  Read those data in:

```{r echo=TRUE, eval=FALSE}
# alternative layout of precipitation data
csv_file <- "/Users/bartlein/Projects/geog495/data/csv/EugeneClim-short-alt-pvars.csv"
eugclim_alt <- read_csv(csv_file)
eugclim_alt
```

> Q2 Describe the different form of the two tables (`eugclim` and `eugclim_alt`).  Can you think of a way to produce a time-series plot of precipitation using the data in `eugclim_alt`?  (If so, show the code for doing that, and if not, why not?)

Now use the `gather()` and `spread()` functions from the `tidyr` package to reshape the data.  This is done here in two steps: 

```{r echo=TRUE, eval=FALSE}
# reshape by gathering and spreading
# 1) gather
eugclim_alt2 <- gather(eugclim_alt, `1`:`12`, key="month", value="cases")
eugclim_alt2$month <- as.integer(eugclim_alt2$month)
eugclim_alt2
```
```{r echo=TRUE, eval=FALSE}
# 2) spread
eugclim_alt3 <- spread(eugclim_alt2, key="param", value=cases)
eugclim_alt3
```

Plot the reshaped data (`eugclim_alt3`) to verify that they indeed have been reshaped correctly.

```{r echo=TRUE, eval=FALSE}
# plot the reshaped data
eugclim_alt3$yrmn <- eugclim_alt3$year + (as.integer(eugclim_alt3$month)-1)/12
plot(eugclim_alt3$prcp ~ eugclim_alt3$yrmn, type="o", pch=16, col="blue", xaxp=c(2013, 2016, 3))
```

> Q3:  Compare `eugclim_alt2` and `eugclim_alt3`.  What did the application of `gather()` do in creating `eugclim_alt2`, and what did the application of `spread()` do in creating `eugclim_alt3`?

> Q4:  What is the benefit of reshaping the data in **R** as opposed to simply doing that in Excel or a text editor?


**4.  A little matrix algebra**

Create three matrices, **A**, **B**, and **C**:

```{r echo=TRUE, eval=FALSE}
# create three matrices
# default fill method: byrow = FALSE
A <- matrix(c(6, 9, 12, 13, 21, 5), nrow=3, ncol=2)
A
```
```{r echo=TRUE, eval=FALSE}
# same elements, but byrow = TRUE
B <-  matrix(c(6, 9, 12, 13, 21, 5), nrow=3, ncol=2, byrow=TRUE)
B
```
```{r echo=TRUE, eval=FALSE}
# a third matrix
C <- matrix(c(1,2,3,4,5,6,7,8,9), nrow=3, ncol=3)
C
```

>Q5 Describe the shapes of the three matrices.  (Note that the `dim()` function applied to a matrix (e.g. `dim(A)`) displays the number of rows and the number of columns in the matrix.)

*Matrix addition*

Add **A** and **B**:

```{r echo=TRUE, eval=FALSE}
# matrix addition
F <- A + B
F
```

Now try adding **A** and **C**:

```{r echo=TRUE, eval=FALSE}
G <- A + C
```

> Q6:  What happend?  Can **A** and **C** be added?  Why not?  (Again, the `dim()` function might be useful.)

*Matrix multiplication*

Matrix multiplication (as distinct from element-by-element multiplication) produces a new matrix whose elements are sums of squares and cross products of the elements of matrices being multiplied 
(see [matrix.pdf](https://pjbartlein.github.io/GeogDataAnalysis/topics/matrix.pdf)).  Matrix multiplication uses `%*%` as the operator. 

"Postmultiply" the matrix **C** by **A**:

```{r echo=TRUE, eval=FALSE}
# matrix multiplication
Q <- C %*% A
Q
```

... and try to postmultiply **A** by **B** (e.g. `T <- A %*% B`) 

>Q7:  What happens here?  What are the dimensions of **C**?  What does the message `non-conformable arguments` imply about the shapes of **A** and **B**?

*Matrix inversion*

To illustrate matrix inversion (i.e. the matrix algebra version of scalar division), a realistic matrix can be used, in this case the correlation matrix of the temperature variables in the `orstationc` data set:

```{r echo=TRUE, eval=FALSE}
# a realistic matrix, orstationc temperature-variable correlation matrix
R <- cor(cbind(orstationc$tjan, orstationc$tjul, orstationc$tann))
R
```

Get the inverse of **R**:

```{r echo=TRUE, eval=FALSE}
# matrix inversion
Rinv <- solve(R)
Rinv
```

One property of the inverse matrix is that when pre- or postmultiplied by the original matrix, the identity matrix, **I** should be produced.

> Q8:  Check to see if `Rinv` is indeed the inverse of `R`.  (Show the results of the check.) 

